Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-19 manuscript.tex
Update on Overleaf.

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is a all-or-none event ---it can occur or not without further details--- at timings asynchronous within a cell assembly, i.e.~it can occur at any continuous time. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states in order to generate actions. These are essential for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of  Neural activity. We sometimes use the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that this neural activity is related to a specific variable, and since Galvani's experiments, we know that electrical activity can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A very important characteristic of neural activity is the fact that it consists mainly of action potentials, or spikes, which are brief (about one millisecond) electrochemical impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing).  The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.

@@

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented on the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``that like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
Why do neurons communicate through spikes? By definition, a spike, or action potential, is an all-or-none event ---it can occur or not without further details--- at asynchronous timings within a cell assembly, i.e.~it can occur at any continuous time. This stands in stark contrast to the discretized timing classically used in digital processing at the base of modern-day neural networks. In the living world, neural systems almost systematically use this so-called event-based representation. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging topic in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal mission in the age of the digital economy and of global warming.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
 Neural activity is directly influenced by our immediate environment and by internal states in order to generate actions. These are essential for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of  Neural activity. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality. In this respect, it is known that this neural activity is related to a specific variable, and since Galvani's experiments, we know that electrical activity can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A very important characteristic of neural activity is the fact that it consists mainly of action potentials, or spikes, which are brief (about one millisecond) electrochemical impulses that propagate along the axons of neurons. These have the particularity of being essentially binary in their amplitude (i.e. they are prototypical, all or nothing).  The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Neurophysiologists typically use firing sequences to characterize the activity of neurons using different statistics on their individual timing~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.  However, computational neuroscience models have suggested that precise timing may also play a crucial role, and that neurons may be as well integrators than synchrony detectors~\citep{paugam-moisy_computing_2012}. We will investigate this very hypothesis in this review.

@@

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented to the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``that like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
